Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-08-27 Documents/Paper/ms.tex
Writing in the Sinewave section

\subsection{Prior Distributions}
In the notation of Section~\ref{sec:introduction}, the parameters of the
each object are
\begin{eqnarray}
\xx_i &=& \{A_i, T_i, \phi_i\}.
\end{eqnarray}
For the interim prior $p(\xx_i | \hyperparams)$, we introduced a single
hyperparameter $\mu$, such that $A_i$ given $\mu$ has an exponential prior
with mean $\mu$. For the periods, we assigned a log-uniform distribution for
the periods between fixed boundaries, and a uniform distribution for the phases
between 0 and $2\pi$. In other words, our prior is only hierarchical for the
amplitudes $\{A_i\}$.

The model for the shape of the (noise-free) signal is
\begin{eqnarray}
y(t) &=& \sum_{i=1}^N A_i \sin \left(\frac{2\pi t}{T_i} + \phi_i\right)
\end{eqnarray}
where there are $N$ sinusoids in the signal, the
amplitudes are $\{A_i\}$, the periods are $\{T_i\}$, and the phases are
$\{\phi_i\}$.
The sampling distribution (probability distribution for the data given the
parameters) is a normal (gaussian) distribution with mean zero and standard
deviation $\sigma$, applied independently to each data point:
\begin{eqnarray}
Y_i | N, \{A_i\}, \{T_i\}, \{\phi_i\}, \sigma \sim
\mathcal{N}\left(\mu(t_i), \sigma^2\right).
\end{eqnarray}
This also depends on an additional noise standard deviation parameter $\sigma$
which will also be inferred. We used a log-uniform prior for $\sigma$ between
$10^{-3}$ and $10^{3}$.

\subsection{Simulated Dataset}
To demonstrate the techniques,
we simulated some data based on the assumption $N=2$, i.e. the true signal was
composed of two sinusoids.
The simulated data is shown in Figure~\ref{fig:sinewave_data}, and shows a
large, low period oscillation with a smaller, much faster oscillation
superimposed. The noise level is such that the fast oscillation is difficult
to detect. The true values of the parameters were
$N=2$, $\mathbf{A} = \{1, 0.3\}$,
$\mathbf{T}=\{30, 2\}$, and $\boldsymbol{\phi} = \{0, 1\}$. The signal was
observed at $n=1001$ points equally spaced between $t=0$ and $t=100$.
The true form of the signal is:
\begin{eqnarray}
y(t) = \sin\left(\frac{2\pi t}{30}\right) +
0.3 \sin\left(\frac{2\pi t}{2} + 1\right)
\end{eqnarray}
and the noise standard deviation was $\sigma = 1$. The data $\{Y_i\}_{i=1}^n$
is displayed in Figure~\ref{fig:sinewave_data}.

@@

\subsection{Results}
We ran DNS on the simulated data shown in Figure~\ref{fig:sinewave_data}.
The shape of the likelihood function, with respect to the prior mass (the
usual likelihood curve associated with Nested Sampling), us shown in
Figure~\ref{fig:sinewaves_likelihood}. Two phase transitions can be seen. The
first, at $\log(X) \approx -10$ nats,
separates ``noise-only'' models (where the
entire dataset is accounted for by the noise term) from ``one-sinusoid'' models
where $N=1$. At $\log(X) \approx -28$ nats, a second phase transition separates
$N=1$ models from $N=2$ models.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{N_result.pdf}
\caption{The inference for $N$, the number of sinusoids, based on the
sinusoid data. The true value was $N=2$.
\label{fig:N_result}}
\end{center}
\end{figure}

@@

>>>>

\subsection{Simulated Dataset}
To demonstrate the techniques,
we simulated some data based on the assumption $N=2$, i.e. the true signal was
composed of two sinusoids.
The simulated data is shown in Figure~\ref{fig:sinewave_data}, and shows a
large, low period oscillation with a smaller, much faster oscillation
superimposed. The noise level is such that the fast oscillation is difficult
to detect. The true values of the parameters were
$N=2$, $\mathbf{A} = \{1, 0.3\}$,
$\mathbf{T}=\{30, 2\}$, and $\boldsymbol{\phi} = \{0, 1\}$. The signal was
observed at $n=1001$ points equally spaced between $t=0$ and $t=100$.
The true form of the signal is:
\begin{eqnarray}
y(t) = \sin\left(\frac{2\pi t}{30}\right) +
0.3 \sin\left(\frac{2\pi t}{2} + 1\right)
\end{eqnarray}
and the noise standard deviation was $\sigma = 1$.

@@

\subsection{Prior Distributions}
To infer $N$, the number of sinusoids, as well as all the periods, amplitudes,
and phases, we need to define prior distributions.
In the notation of Section~\ref{sec:introduction}, the parameters of the
each object are
\begin{eqnarray}
\xx_i &=& \{A_i, T_i, \phi_i\}.
\end{eqnarray}
For the interim prior $p(\xx_i | \hyperparams)$, we introduced a single
hyperparameter $\mu$, such that $A_i$ given $\mu$ has an exponential prior
with mean $\mu$. For the periods, we assigned a log-uniform distribution for
the periods between fixed boundaries, and a uniform distribution for the phases
between 0 and $2\pi$. In other words, our prior is only hierarchical for the
amplitudes $\{A_i\}$. The prior for $\mu$ was a log-uniform prior, with
probability density $\propto 1/\mu$, where $\mu \in [10^{-3}, 10^3]$.

The model for the shape of the (noise-free) signal is
\begin{eqnarray}
y(t) &=& \sum_{i=1}^N A_i \sin \left(\frac{2\pi t}{T_i} + \phi_i\right)
\end{eqnarray}
where there are $N$ sinusoids in the signal, the
amplitudes are $\{A_i\}$, the periods are $\{T_i\}$, and the phases are
$\{\phi_i\}$.
The sampling distribution (probability distribution for the data given the
parameters) was a normal (gaussian) distribution with mean zero and standard
deviation $\sigma$, applied independently to each data point:
\begin{eqnarray}
Y_i | N, \{A_i\}, \{T_i\}, \{\phi_i\}, \sigma \sim
\mathcal{N}\left(\mu(t_i), \sigma^2\right).
\end{eqnarray}
This also depends on an additional noise standard deviation parameter $\sigma$
which will also be inferred. We used a log-uniform prior for $\sigma$ between
$10^{-3}$ and $10^{3}$.

\subsection{Results}
We ran DNS on the simulated data shown in Figure~\ref{fig:sinewave_data}.
The shape of the likelihood function with respect to the prior mass (the
usual likelihood curve associated with Nested Sampling) is shown in
Figure~\ref{fig:sinewaves_likelihood}. Two phase transitions can be seen,
where this curve is concave up. The
first phase transition, at $\log(X) \approx -10$ nats,
separates ``noise-only'' models (where the
entire dataset is accounted for by the noise term) from ``one-sinusoid'' models
where $N=1$. At $\log(X) \approx -28$ nats, a second phase transition separates
$N=1$ models from $N=2$ models. This phase transition causes a double-peaked
feature in the posterior weights (lower panel of
Figure~\ref{fig:sinewaves_likelihood}).

@@

Although DNS explores a distribution other than the posterior (a mixture of
constrained priors is used instead), the saved samples can be assigned
importance weights so we can represent the posterior distribution. The
marginal posterior distribution for $N$ is shown in Figure~\ref{fig:N_result}.
The most probable value is $N=2$ which is also the true value, and there is
a small probability for $N=1$. The decreasing probabilities for $N > 2$ are
due to the well-known natural ``Occam's Razor'' effect that can occur in
Bayesian Inference \citep[e.g.][Chapter 28]{mackay}.

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{N_result.pdf}
\caption{The inference for $N$, the number of sinusoids, based on the
sinusoid data. The true value was $N=2$, which is also the most probable value
given the data. Of course, this is not always the case.
\label{fig:N_result}}
\end{center}
\end{figure}

@@

\bibitem[MacKay(2003)]{mackay} MacKay, David J.~C., 2003, ``Information theory, inference, and learning algorithms''. Vol. 7.
Cambridge: Cambridge university press, 2003.
